Automated Verification for Real-Time Systems using Implicit Clocks and an Extended Antimirov Algorithm

### 2022-04-05 paper/tv.tex
m

We investigate an alternative approach for modeling and verifying 
compositional real-time systems. In this work, 
we propose a novel temporal specification language, 
which enables a compositional verification via a  Hoare-style 
forward verifier and a term rewriting system (TRS). 
More specifically, we specify system behaviours in the form of 
{Timed Effects} (\timedEffects), which integrates the Kleene Algebra with dependent values and arithmetic constraints, 
%gaining the expressive power beyond finite-state machines; and (ii) introduces a new parallel operator 
to 
%capture the parallelism between event-triggered (sync) and  time-triggered (async) execution traces, 
provide real-time abstractions into traditional linear temporal logics. 
%Resource usages, provided by a \emph{dynamic ticks}\cite{von2017real}, incorporating the time-triggered (physical time) and event-triggered (logical time) execution loops. 
For example, one safety property, \textit{"The event \anyevent{Done} 
will be triggered no later than one time unit"}\footnote{Without loss of generality, 
we use integer values to represent time units in this 
paper, while it can bes extended to real numbers and other 
time measurement units.}, is expressed in \timedEffects\ as: 
\begin{align*}
&\code{\effect \defeq \  0 {\leq} t {<}1, ({\_^\star} \cdot \anyevent{Done} ) \# t}.
\end{align*}  
%The timed effects incorporates different kinds of existing temporal logics, such as linear-time temporal logic (LTL)
%and .
%which replaces temporal operators by time-constrained versions from 
Here, \code{\#} is a novel operator specifying the \emph{real-time} 
constraints for the \emph{logical-time} sequences \cite{von2017real}; 
\code{\_} is a wildcard matching to any event; 
Kleene star \code{\star} denotes trace repetition.
The above formula \code{\effect} corresponds to `\code{\Diamond_{[0, 1)}\ }\anyevent{Done}' 
in metric temporal logic (MTL), reads \textit{"within one second, 
\anyevent{Done} finally happens"}. Moreover, the time bounds 
can be dependent on the program inputs. For example, as the program 
shown in \autoref{fig:Value_dependent_intro}, we mark the precondition and postcondition
using keywords \code{\emph{require}} and \code{\emph{ensure}} respectively.

@@

\begin{wrapfigure}{L}{0.4\columnwidth}
  \vspace{-2mm}
\begin{lstlisting}
void addSugar (int n) 
/*require n>=0 , _^*
  ensure  t>=n, Sugar#t */
{ if (n == 0) { 
    event ["Sugar"];} 
  else {
    timeout (oneSugar() , 1);
    addSugar (n-1);}} 
\end{lstlisting}
\vspace{-1mm}
\caption{Value-dependent specification.} 
\label{fig:Value_dependent_intro}
\vspace{-1mm}
\end{wrapfigure}

The function 
\code{\emph{addSugar}} is intended to takes a parameter \codem{n}, representing 
the portion of the sugar we need to add. When \codem{n} decreases to \codem{0}, 
it simply raises an event \anyevent{Sugar} to mark the end of the process. 
Otherwise, it adds one portion of the sugar then recursively calls \code{\emph{addSugar}} with parameter \codem{n-1}. 
Statement \codem{timeout(e, d)} sets a timer which executes code \codem{e} once the timer expires at time \codem{d}. 
Therefore, the time spent on adding one portion of the sugar is no less than 1 time unit.

@@

Here, \code{c} is for integer constants, \code{b} is for boolean values, 
 \code{mn} and \code{x} are for meta-variables.  
 \textbf{var} represents the countably infinite set of arbitrary distinct identifiers. 
\code{ev} refers to a singleton event coming from the finite 
set of events \code{\Sigma}. 
%Without loss of generality, any event can be further parametrized with 
%constant values, e.g. \code{\anyevent{a}.0} and  \code{\anyevent{a}}.1 are distinct two events. 
%leading to infinite alphabet.
We assume that programs we use are well-typed conforming to basic types \code{\tau} (we take () as the \code{void} type).
%Each events raising expression \code{\textbf{event}} is parametrized with an event, to trigger such a single event. 
A program \code{\mathcal{P}} comprises a list of 
global values initializations \code{(x:=v)^*} and 
a list of 
method declarations \code{{meth^*}}. 
Here, we use the \code{*} superscript to denote a finite list (possibly empty) of items, for example, \code{{x^*}} refers to a list of variables, \code{x_1, ...,\ x_n}. 
%Besides, \code{d} denotes the name of a user-defined data type, \code{f} denotes a field name.

@@

Expressions comprise simple values \code{v};
a guarded process is written as \code{[b]e} where if b is true, 
then it behaves as \code{e}, else it idles until \code{b} becomes true; 
%one-level field access \code{v.f} (rather than \code{v.f1.f2...} for the sake of simplicity), 
%data structure allocation \code{\textbf{new}\ d({v^*})},  
%local variable declaration \code{\tau\ x;\ e}, 
method calls \code{mn({x^*})}; 
%variable assignments \code{x{:=}e}; 
%field update \code{v.f:=e}, 
expression sequences \code{e_1;e_2};
parallel composition \code{e_1 || e_2}, where \code{e_1} and \code{e_2} 
may communicate via multi-party event synchronization or shared variables; 
conditional expressions \code{{if}\ v\ {then}\ e_1\ {else}\ e_2}. 
The expression \code{ \textbf{event} [\anyevent{A}(v)]\{(x:=v)^*\} } 
raises an event \anyevent{A} parametrized with \code{v}, and takes a set of 
assignment actions \code{(x:=v)^*}.

@@

%     \m{(Types)} &  \code{t}&\code{  ::= }&
% \code{d\ |\ \tau} 
% \\    
  \m{(Method\ Def.)} &  \code{meth}&\code{  ::= }&
  \code{ mn  \ {(x)^*}\ \{\textbf{requires}\ \effect_{pre}\ \textbf{ensures}\  \effect_{post}\} \ \{ e \}}
  \\
  \m{(Values)} & \code{v} &\code{  ::= } & ()\ |\ c \ | \ b \ |\ x\  
  \\
  \m{(Expressions)} &  \code{e}&\code{  ::= }&
    \code{ v \ | \ 
    [b]\ e \ |\ mn({v^*}) \ }
  \code{| 
 % v.f:=e \ | 
 \ e_1;e_2\ | \ e_1 || e_2 \ 
 % |\ {if}\ b\ {then}\ e_1\ {else}\ e_2 \
 |\ \textbf{event} [\anyevent{A}(v, \alpha^*)] 
 }  \\

@@

    \multicolumn{4}{c}{
 % \code{k^\tau:constant\ of\ type\ \tau\ } 
 % \qquad\quad
\code{c  \in \mathbb{Z}}
\qquad \ \ 
\code{b  \in \mathbb{B}}
\qquad \ \ 
\code{x, mn  ::\in \textbf{var}}  
\qquad\ \ 
 \code{(Action\ labels)\ \anyevent{A}  \in \Sigma 
\qquad\ \ 
\m{(Time\ Bounds)}\ \code{d} \code{\in \mathbb{N}}    
}   
    }\\  
    \hline
  \end{array}  
  $
 \caption{A Core First-Order Timed Imperative Language, \timedL.} 
 \label{fig:code_language}

@@

Furthermore, a number of timed process constructs can be used to capture common 
real-time system behavior patterns. Without loss of generality, we 
assume d is a non-negative integer constant.  $\textbf{delay} [d]$ idles for exactly $d$ 
time units. In process $e_1\ \textbf{timeout} [d]\ e_2$, the first observable event of 
$e_1$ shall occur before d time units elapse (since the process starts). 
Otherwise, $e_2$ takes over control after exactly $d$ time units elapse. 
Process $\textbf{deadline} [d]\ e$ constrains $e$ to terminate before $d$ time units. 
In this setting, clock variables are 
made implicit and hence they cannot be compared with each other 
directly, which potentially allows efficient clock manipulation and 
hence system verification.

@@

A transition of the system is of the form $c \xrightarrow[]{\text{x}} c^\prime$ where $c$ and $c^\prime$ are 
the system configurations before and after the transition respectively. 
We adopt the following naming convention for transition labels: 
$t$ denotes a non-negative real number; $\tau$ denotes an invisible event;
For instance, $c \xrightarrow[]{\text{t}} c^\prime$ denotes a transition of $t$ time units elapsing. 
In the following, we present the firing rules which are associated with 
the timed process constructs. 
\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   Wait                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  t {\leq} d 
}{(V, \textbf{delay}\ [d]) \xrightarrow[]{\text{t}} (V, \textbf{delay}\ [d \text{-}t])}\ [de1]} 
\qquad \qquad 
\code{\frac{
}{(V, \textbf{delay}\ [0]) \xrightarrow[]{\tau} (V, ())}\ [de2]} 
\end{flalign*}

\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   timeout                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  (V,e_1) \xrightarrow[]{\text{e}} (V^\prime,e_1^\prime)
}{(V, e_1\ \textbf{timeout} \  [d] \ e_2) \xrightarrow[]{\text{e}} (V^\prime,e^\prime)}\ [to1]} 
\quad  
\code{\frac{
  (V,e) \xrightarrow[]{\tau} (V^\prime,e^\prime)  \quad (t {\leq} d)
}{(V, e_1\ \textbf{timeout} \  [d] \ e_2) \xrightarrow[]{\text{t}} (V, \textbf{timeout}\ e^\prime \ [d \text{-}t])}\ [to2]} 
\\
\code{\frac{
  (V,e_1) \xrightarrow[]{\text{t}} (V,e_1^\prime)  \quad (t {\leq} d) 
}{(V, e_1\ \textbf{timeout} \ [d] \ e_2) \xrightarrow[]{\text{t}} (V,e_1^\prime\ \textbf{timeout} \  [d\text{-}t] \ e_2)}\ [to3]} 
\quad  
\code{\frac{
}{(V, e_1\ \textbf{timeout} \  [0] \ e_2) \xrightarrow[]{\tau} (V, e_2)}\ [to4]} 
\end{flalign*}

\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   deadline                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  (V,e) \xrightarrow[]{\text{x}} (V^\prime,e^\prime)
}{(V, \textbf{deadline}\  [d] \ e ) \xrightarrow[]{\text{x}} (V^\prime, \textbf{deadline} \ [d] \ e^\prime)}\ [dl1]} 
\quad  
\code{\frac{
  (V,e) \xrightarrow[]{\text{t}} (V,e^\prime)  \quad (t {\leq} d)
}{(V, \textbf{deadline} \  [d]\ e ) \xrightarrow[]{\text{t}} (V, \textbf{deadline} \ [d \text{-}t] \ e^\prime )}\ [dl2]} 
\end{flalign*}

>>>>

We investigate an alternative approach for modeling and verifying 
compositional real-time systems. In this work, 
we propose a novel temporal specification language, 
which enables a compositional verification via a  Hoare-style 
forward verifier and a term rewriting system (TRS). 
More specifically, we specify system behaviours in the form of 
{Timed Effects} (\timedEffects), which integrates the Kleene Algebra with dependent values and arithmetic constraints, 
%gaining the expressive power beyond finite-state machines; and (ii) introduces a new parallel operator 
to 
%capture the parallelism between event-triggered (sync) and  time-triggered (async) execution traces, 
provide real-time abstractions into traditional linear temporal logics. 
%Resource usages, provided by a \emph{dynamic ticks}\cite{von2017real}, incorporating the time-triggered (physical time) and event-triggered (logical time) execution loops. 
For example, one safety property, \textit{"The event \anyevent{Done} 
will be triggered no later than one time unit"}\footnote{Without loss of generality, 
we use integer values to represent time units in this 
paper, while it can bes extended to real numbers and other 
time measurement units.}, is expressed in \timedEffects\ as: 
\code{\effect \defeq \  0 {\leq} t {<}1, ({\_^\star} \cdot \anyevent{Done} ) \# t}.
%\begin{align*}
%&\end{align*}  
%The timed effects incorporates different kinds of existing temporal logics, such as linear-time temporal logic (LTL)
%and .
%which replaces temporal operators by time-constrained versions from 
Here, \code{\#} is a novel operator specifying the \emph{real-time} 
constraints for the \emph{logical-time} sequences \cite{von2017real}; 
\code{\_} is a wildcard matching to any event; 
Kleene star \code{\star} denotes trace repetition.
The above formula \code{\effect} corresponds to `\code{\Diamond_{[0, 1)}\ }\anyevent{Done}' 
in metric temporal logic (MTL), reads \textit{"within one second, 
\anyevent{Done} finally happens"}. Moreover, the time bounds 
can be dependent on the program inputs. For example, as the program 
shown in \autoref{fig:Value_dependent_intro}, we mark the precondition and postcondition
using keywords \code{\emph{require}} and \code{\emph{ensure}} respectively.

@@

\begin{wrapfigure}{L}{0.39\columnwidth}
  \vspace{-2mm}
\begin{lstlisting}
void addSugar (int n) 
/*require n>=0, _^*
  ensure  t>=n, Sugar#t */
{ if (n == 0) { 
    event ["Sugar"];} 
  else {
    timeout (oneSugar() , 1);
    addSugar (n-1);}} 
\end{lstlisting}
\vspace{-1mm}
\caption{Value-dependent specification.} 
\label{fig:Value_dependent_intro}
\vspace{-1mm}
\end{wrapfigure}

The function 
\code{\emph{addSugar}} is intended to takes a parameter \codem{n}, representing 
the portion of the sugar we need to add. When \codem{n} decreases to \codem{0}, 
it simply raises an event \anyevent{Sugar} to mark the end of the process. 
Otherwise, it adds one portion of the sugar then recursively calls \code{\emph{addSugar}} with parameter \codem{n-1}. 
The use of statement \codem{timeout(e, d)} is standard, 
which sets a timer which executes code \codem{e} once the timer 
expires at time \codem{d}. 
Therefore, the time spent on adding one portion of the sugar is 
no less than 1 time unit.

@@

Here, \code{c} is for integer constants, \code{b} is for boolean values, 
 \code{mn} and \code{x} are for meta-variables.  
 \textbf{var} represents the countably infinite set of arbitrary distinct identifiers. 
\code{ev} refers to a singleton event coming from the finite 
set of events \code{\Sigma}. 
%Without loss of generality, any event can be further parametrized with 
%constant values, e.g. \code{\anyevent{a}.0} and  \code{\anyevent{a}}.1 are distinct two events. 
%leading to infinite alphabet.
%We assume that programs we use are well-typed conforming to basic types \code{\tau} (we take () as the \code{void} type).
%Each events raising expression \code{\textbf{event}} is parametrized with an event, to trigger such a single event. 
A program \code{\mathcal{P}} comprises a list of 
global values initializations \code{\alpha^*} and 
a list of 
method declarations \code{{meth^*}}. 
Here, we use the \code{*} superscript to denote a finite list (possibly empty) of items, for example, \code{{x^*}} refers to a list of variables, \code{x_1, ...,\ x_n}. 
%Besides, \code{d} denotes the name of a user-defined data type, \code{f} denotes a field name.

@@

Expressions comprise simple values \code{v};
a guarded process is written as \code{[b]e} where if b is true, 
then it behaves as \code{e}, else it idles until \code{b} becomes true; 
%one-level field access \code{v.f} (rather than \code{v.f1.f2...} for the sake of simplicity), 
%data structure allocation \code{\textbf{new}\ d({v^*})},  
%local variable declaration \code{\alpha; e}, 
method calls \code{mn({x^*})}; 
variable assignments \code{\alpha}; 
%field update \code{v.f:=e}, 
expression sequences \code{e_1;e_2};
parallel composition \code{e_1 || e_2}, where \code{e_1} and \code{e_2} 
may communicate via multi-party event synchronization or shared variables. 
%conditional expressions \code{{if}\ v\ {then}\ e_1\ {else}\ e_2}. 
The expression \code{ \textbf{event} [\anyevent{A}(v)] } 
raises an event \anyevent{A} parametrized with \code{v}. 
%and takes a set of 
%assignment actions \code{(x:=v)^*}. 
Furthermore, a number of timed process constructs can be used to capture common 
real-time system behavior patterns. Without loss of generality, we 
assume d is a non-negative integer constant.  $\textbf{delay} [d]$ idles for exactly $d$ 
time units. In process $e_1\ \textbf{timeout} [d]\ e_2$, the first observable event of 
$e_1$ shall occur before d time units elapse (since the process starts). 
Otherwise, $e_2$ takes over control after exactly $d$ time units elapse. 
Process $\textbf{deadline} [d]\ e$ constrains $e$ to terminate before $d$ time units.

@@

%     \m{(Types)} &  \code{t}&\code{  ::= }&
% \code{d\ |\ \tau} 
% \\    
  \m{(Method\ Def.)} &  \code{meth}&\code{  ::= }&
  \code{ mn  \ {(x)^*}\ \{\textbf{requires}\ \effect_{pre}\ \textbf{ensures}\  \effect_{post}\} \ \{ e \}}
  \\
  \m{(Values)} & \code{v} &\code{  ::= } & ()\ |\ c \ | \ b \ |\ x\  
  \\
  \m{(Expressions)} &  \code{e}&\code{  ::= }&
    \code{ v \ | \ 
    \alpha  \ | \ 
    [b]\ e \ |\ mn({v^*}) \ }
  \code{| 
 % v.f:=e \ | 
 \ e_1;e_2\ | \ e_1 || e_2 \ 
 % |\ {if}\ b\ {then}\ e_1\ {else}\ e_2 \
 |\ \textbf{event} [\anyevent{A}(v)] 
 }  \\

@@

    \multicolumn{4}{c}{
 % \code{k^\tau:constant\ of\ type\ \tau\ } 
 % \qquad\quad
\code{c  \in \mathbb{Z}}
\qquad \ \ 
\code{b  \in \mathbb{B}}
\qquad \ \ 
\code{x, mn  ::\in \textbf{var}}  
\qquad\ \ 
 \code{(Action\ labels)\ \anyevent{A}  \in \Sigma 
\qquad\ \ 
\m{(Time\ Bounds)}\ \code{d} \code{\in \mathbb{N}}    
}   
    }\\  
    \hline
  \end{array}  
  $
 \caption{A core first-order imperative language with timed constructs, \timedL.} 
 \label{fig:code_language}

@@

We remark additional process constructs (e.g. if-then-else, while, etc.) 
can be defined using the above. In this setting, clock variables are 
made implicit and hence they cannot be compared with each other 
directly, which potentially allows efficient clock manipulation and 
hence system verification.

@@

A transition of the system is of the form $c \xrightarrow[]{l} c^\prime$ where $c$ and $c^\prime$ are 
the system configurations before and after the transition respectively. 
Transition labels \code{l} consist: 
%We adopt the following naming convention for transition labels: 
$t$, denoting a non-negative real integer; 
$\tau$, denoting an invisible event;
$\anyevent{A} \in \Sigma $ is an observable 
event; and $\sigma \in \Sigma\cup \{\tau\}$. 
For instance, $c \xrightarrow[]{\text{t}} c^\prime$ denotes a transition of \codem{t} time units elapsing. 
In the following, we present the firing rules which are associated with 
the timed process constructs.

\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   Wait                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  t {\leq} d 
}{(V, \textbf{delay}[d]) \xrightarrow[]{\text{t}} (V, \textbf{delay} [d \text{-}t])}\ [de_1]} 
\qquad \qquad 
\code{\frac{
}{(V, \textbf{delay} [0]) \xrightarrow[]{\tau} (V, ())}\ [de_2]} 
\end{flalign*}

The above captures behaviors of process \code{\textbf{delay}[d]}. 
Rule \code{de_1} states that the process may idle for 
any amount of time as long as it is less than or equal to \codem{d} time units; 
Rule \code{de_2} states that the process terminates immediately after 
\codem{d} becomes \codem{0}.
\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   timeout                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  (V,e_1) \xrightarrow[]{\anyevent{A}} (V^\prime,e_1^\prime)
}{(V, e_1\ \textbf{timeout}  [d] \ e_2) \xrightarrow[]{\anyevent{A}} (V^\prime,e^\prime)}\ [to_1]} 
\quad  
\code{\frac{
  (V,e_1) \xrightarrow[]{\tau} (V^\prime,e_1^\prime)  
}{(V, e_1\ \textbf{timeout}  [d] \ e_2) \xrightarrow[]{\text{t}} (V^\prime, e_1^\prime\ \textbf{timeout}[d]\ e_2 )}\ [to_2]} 
\\
\code{\frac{
  (V,e_1) \xrightarrow[]{\text{t}} (V,e_1^\prime)  \quad (t {\leq} d) 
}{(V, e_1\ \textbf{timeout}  [d] \ e_2) \xrightarrow[]{\text{t}} (V,e_1^\prime\ \textbf{timeout} \  [d\text{-}t] \ e_2)}\ [to_3]} 
\quad  
\code{\frac{
}{(V, e_1\ \textbf{timeout}   [0] \ e_2) \xrightarrow[]{\tau} (V, e_2)}\ [to_4]} 
\end{flalign*}
If an observable event \anyevent{A} can be engaged by \code{e_1}, 
then \code{e_1\ \textbf{timeout}[d]\ e_2} becomes \code{e_1^\prime} (rule \code{[to_1]}). 
An invisible transition does 
not solve the choice (rule \code{[to_2]}). 
If \code{e_1} may idle for less than or equal 
to \codem{d} time units, so is the composition (rule \code{[to_3]}). 
When \codem{d} becomes \codem{0}, 
\code{e_2} takes over control by a silent transition (rule \code{[to_4]}).
\begin{flalign*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   deadline                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\code{\frac{
  (V,e) \xrightarrow[]{\anyevent{A}} (V^\prime,e^\prime)
}{(V, \textbf{deadline} [d] \ e ) \xrightarrow[]{\anyevent{A}} (V^\prime, \textbf{deadline}  [d] \ e^\prime)}\ [dl_1]} 
\  
\code{\frac{
  (V,e) \xrightarrow[]{\text{t}} (V,e^\prime)  \quad (t {\leq} d)
}{(V, \textbf{deadline}  [d]\ e ) \xrightarrow[]{\text{t}} (V, \textbf{deadline}  [d \text{-}t] \ e^\prime )}\ [dl_2]} 
\end{flalign*}

Intuitively, \code{\textbf{deadline}\  [d] \ e} behaves exactly as 
\code{e} except that it must terminate before \codem{d} time units.
